---
title: "Simulate Multi-State"
author: "analyze.eng@gmail.com"
date: "2024-06-20"
output: html_document
---

```{r setup, eval= TRUE, echo=FALSE}
## Clear current session
rm(list = ls(all.names=TRUE))

## Free memory, force garbage collection
invisible(gc())

## Main Knitr options
knitr::opts_chunk$set(echo = TRUE)

## Default behavior: Suppress warnings
knitr::opts_chunk$set(eval=TRUE,
                      message=FALSE,warning=FALSE,
                      comment.char=" ")                     

## Print graphics after code chunk
knitr::opts_chunk$set(fig.align="center",
                      fig.show ="asis")

## KHE standard sizing
knitr::opts_chunk$set(fig.width=12,fig.height=6)
```

```{r more_setup}
library(survival)
library(gss)
library(muhaz)
library(umap)

## Set colors from Kevin's favorite subway lines
subway<-c("#0039A6","#EE352E","#FCCC0A","#00933C",
          "#FF6319","#B933AD","#00B2A9","#00A1DE")
palette(subway)
```

# Context

I am interested in the idea that we can find biomarkers of a particular type by studying the joint (OS, PFS) process.  A technical innovation is to consider the multi-state survival process of a particular topology.  

My plan is to generate data following `coxph(OS~b1*X)` and `coxph(PFS~b2*X)` and then to estimate the component hazards.  I think that we will see a peculiar behavior as a function of covariates `(b1, b2)`.  

Some questions about the censoring process are also relevant.  Consider the independent censoring time that may follow `coxph(CENS~b3*X)` and the implications for the component hazards.

Analysis will follow https://cran.r-project.org/web/packages/survival/vignettes/compete.pdf

# TCGA Data Analysis
Data from TCGA-OV is available to show that this data structure exists in reality.  


## Load Data
We are after the general `Surv(tstart,tstop,event)` data structure where `levels(event) = c("ANED","DEATH","PROGRESSION")`.  

```{r}
## Each row is a person
x<-read.table(
            file = "Data/data_clinical_patient.txt",
            sep="\t",
            skip=4,
            row.names=1,
            header=TRUE)

## Convert to clear variables
dat1<-x[,c("OS_STATUS","OS_MONTHS","PFS_STATUS","PFS_MONTHS")]
dat1$time1<-dat1$OS_MONTHS
dat1$event1<-dat1$OS_STATUS=="1:DECEASED"
dat1$time2<-dat1$PFS_MONTHS
dat1$event2<-dat1$PFS_STATUS=="1:PROGRESSION"

## Drop bad records
dat1<-dat1[which(!is.na(dat1$time1)),]

## Construct the gap time
dat1$gaptime<-dat1$time1-dat1$time2
dat1$id<-rownames(dat1)

## Logic
## 1. if there is a gap time, then (0,time2,event="PROG")
##       and we need (time2,time1,event=event1)
## 2. if not then 
##       we have just (0,time1,event=event1)
gate<-interaction(ifelse(dat1$gaptime>0,"gap","nogap"),
                  ifelse(dat1$event1,"died","alive"))
table(gate,useNA="a")

```

We can now start constructing the event-process data structure

```{r}
dat2<-split(dat1,f=gate)

## Paste together the (tstart,tend,event)
## Data structure using rules from above.
dat3<-
rbind(
  data.frame(tstart=0,
             tend=dat2[["nogap.alive"]][,c("time1")],
             event="cens",
             id=dat2[["nogap.alive"]][,"id"]),
  data.frame(tstart=0,
             tend=dat2[["nogap.died"]][,"time1"],
             event="died",
             id=dat2[["nogap.died"]][,"id"]),
  
  data.frame(tstart=0,
             tend=dat2[["gap.died"]][,"time2"],
             event="prog",
             id=dat2[["gap.died"]][,"id"]),
  data.frame(tstart=dat2[["gap.died"]][,"time2"],
             tend=dat2[["gap.died"]][,"time1"],
             event="died",
             id=dat2[["gap.died"]][,"id"]),
  
  data.frame(tstart=0,
             tend=dat2[["gap.alive"]][,"time2"],
             event="prog",
             id=dat2[["gap.alive"]][,"id"]),
  data.frame(tstart=dat2[["gap.alive"]][,"time2"],
             tend=dat2[["gap.alive"]][,"time1"],
             event="cens",
             id=dat2[["gap.alive"]][,"id"])
)
dat3$event<-factor(dat3$event,c("cens","prog","died"))
```

## Look at multistate data

This data structure is amenable to multi-state regression analysis. Here are some of the usual operations. 

```{r}
library(survival)
mfit1<-survfit(Surv(tstart, tend, event) ~ 1, data=dat3, id=id)
print(mfit1,digits=3,rmean=180)

mfit1$transitions

plot(mfit1[,"prog"],xlab="Months After Surgery",
     ylab="Fraction in State",col=1,ylim=c(0,1))
lines(mfit1[,"died"],col=2)
lines(mfit1[,"(s0)"],col=3)
legend("right",text.col=1:3,
       legend=c("Progressed","Dead","ANED"),bty="n")
```

Okay this is a nice way to look at the data.

We should start putting covariates into the problem.

## TCGA covariates

```{r}
x<-read.table(
            file = "Data/data_mrna_seq_v2_rsem.txt",
            sep="\t",
            header=TRUE,
            colClasses = c("character",
                           "character",
                           rep("numeric",300)))
x<-x[x[,1]!="",]     # Drop missing HUGO
bar<-x[,1]  

# Take sum over RSEM expected counts
# this just piles different gene concepts 
# into the same bin
foo<-apply((x[,-(1:2)]),2,
           function(a)tapply(a,bar,sum))
colnames(foo)<-substr(gsub("\\.","-",colnames(foo)),1,12)

## We'll analyze on log
foo<-log(foo)        # some expected counts are <0
foo[is.nan(foo)]<-NA # kill them

sna<-apply(foo,1,var,na.rm=TRUE)
foo<-foo[sna>0,] # Genes have to have some variation

set<-intersect(rownames(dat1),colnames(foo))
dat<-dat1[set,]
rna<-foo[,set]

## drop any with no counts
rna<-rna[!apply(is.na(rna),1,all),]

```

Establish what b1 and b2 are for OS and PFS.

Then establish what they look like for each of the processes.
